### 2022-04-18 paper/localpsf_old3.tex
about to remove old stuff from paper

% REQUIRED
\begin{abstract}
	We present a fast matrix-free method for computing hierarchical matrix (H-matrix) approximations of operators that have locally supported non-negative integral kernels. Such operators arise, for example, as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained inverse problems. H-matrix approximations are highly desirable because there are fast H-matrix methods to add, multiply, invert, and factorize H-matrices, compute matrix-vector products, solve linear systems, and perform other useful matrix computations. Classical H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. Our method computes impulse responses of the operator at a collection of scattered points, then interpolates these known impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known, which allows us to construct the H-matrix. Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. 
	We apply the method to approximate Hessians in large-scale PDE-constrained inverse problems with highly informative data. Numerical results demonstrate that our method substantially outperforms existing state-of-the-art Hessian approximation methods which are based on low-rank approximation. We are able to form high quality approximations of high rank operators using only a small number of operator applies.
\end{abstract}

@@

We present a fast \emph{matrix free} method for approximating operators, $\Aop:L^2(\Omega) \rightarrow L^2(\Omega)'$, that have locally supported non-negative integral kernels. Here $\Omega \subset \mathbb{R}^d$ is a bounded domain, $d$ is the spatial dimension, and $L^2(\Omega)'$ is the space of continuous linear functionals on $L^2(\Omega)$ (i.e., the topological dual space). Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{le1997non,saad1999distributed,TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps) \cite{DTNMAP}, covariance operators in spatial statistics \cite{GeogaEtAl20,ChenStein21,LindgrenRueLindstrom11}, blurring operators in imaging \cite{DenisEtAl11,NagyOleary98}, and Hessians in distributed parameter PDE-constrained optimization and inverse problems \cite{AlgerEtAl19,IsaacEtAl15}.

By ``matrix free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and its transpose\footnote{Recall that the transpose $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(v) = \left(\Aop v\right)(u)$ for all $u,v \in L^2(\Omega)$, where the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(v)$ is the result of applying that linear functional to $v \in L^2(\Omega)$, and similar for the analogous operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other costly computational procedure. 
%After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representations of $\Aop$ and $\Aop^T$, but cannot easily compute individual matrix entries.

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to the inverse problem, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
%In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. 
%More accessible Hessian approximations are highly desirable because they permit fast solution of the minimization problem via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for the parameter \cite{LAPLACEAPPROXIMATION}. 
The most popular matrix-free Hessian approximation methods are based on low rank approximation \cite{BuiEtAl13,CuiEtAl14,FlathEtAl11,PetraEtAl14,SpantiniEtAl15} because there are efficient methods for constructing low rank approximations of a matrix using only matrix-vector products \cite{Cheng05,HalkoMartinssonTropp11}. 
%This includes the Lanczos method, the randomized singular value decomposition \cite{HMTRANDOM}, and CUR approximation/skeletonization \cite{CUR,MAHONEY}. 
However, low rank approximation methods suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator applies are required to form an accurate ``low rank'' approximation \cite{Alger19,AlgerEtAl17}. High-rank matrix free approximation methods are needed.

\begin{figure}
	\begin{center}
		\includegraphics[scale=0.4]{bct1.eps}
	\end{center}
	\caption{Hierarchical matrix block structure for the Stokes Hessian. The red blocks (tiny) are dense. All other blocks are low rank.
	}
	\label{fig:hmatrix_intro}
\end{figure}

Our method forms a hierarchical matrix (H-matrix) \cite{Hackbusch99,} approximation of a discretized version of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank, even though the matrix as a whole may be high-rank (see Figure \ref{fig:hmatrix_intro} and Appendix \ref{app:h_matrix}). H-matrix methods permit us to perform matrix-vector products cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving the H-matrix that cannot easily be done using the original operator. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. Typically, the work and memory required to perform these operations for an $N \times N$ H-matrix scales as $O(r^2 N \log(N)^a)$, where $r$ is the largest rank of the off-diagonal blocks, and $a \in \{0,1,2,3\}$. The exact cost depends on the type of H-matrix used and operation performed; for more details see \cite{GrasedyckHackbusch03,Hackbusch15}. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch12.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch19.png}
	\end{subfigure}
	\caption{Batches of normalized impulse responses for the Stokes inverse problem Hessian. Left: 2nd batch. Mid: 13th batch. Right: 19th batch. Brightness scale differs between the subfigures. Black stars indicate the sample point locations at which the point source is applied. Dashed gray ellipses indicate the estimated support ellipsoids. The large circle is the boundary of the domain. 
		%		Early batches tend to have lots of ellipsoids near the boundary, as compared to later batches. 
		%		This is because our greedy ellipsoid packing algorithm tries to pick sample points that are far away from sample points in previously chosen batches.
	}
	\label{fig:batches_intro}
\end{figure}

%Since our end goal is an H-matrix, it is natural to ask why we do not build the H-matrix directly using classical H-matrix construction methods. The reason is that 
Classical H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available matrix free. Recently there have been improvements in matrix-free H-matrix construction methods based on the so-called ``peeling process'' \cite{LinYing11,Martinsson11,Martinsson16,MartinssonTropp20}, which we do not use here. 
These alternative methods have, been applied to form H-matrix representations of Hessians in PDE constrained inverse problems \cite{AmbartsumyanEtAl20,HartlandEtAl22}. 
While methods based on the peeling process are asymptotically scalable in theory, in practice the required number of matrix vector products is large. Matrix-free H-matrix construction methods are currently a topic of considerable research, so future algorithmic improvements may make matrix-free H-matrix methods based on the peeling process more efficient in the future.

We construct the H-matrix using \emph{impulse response interpolation}. The impulse response, $\phi_x$, associated with a point $x$ is the Riesz representation\footnote{Recall that the Riesz representative of a functional $\rho \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\rho^* \in L^2(\Omega)$ such that $\rho(v) = \left(\rho^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} of the linear functional that results from applying $\Aop$ to a delta distribution (i.e., point source, impulse) centered at $x$. We compute batches of impulse responses by applying $\Aop$ to weighted sums of delta distributions associated with collections of points scattered throughout the domain (see Figure \ref{fig:batches_intro}). Then we interpolate shifted versions of these impulse responses to approximate arbitrary entries of the operator's integral kernel. This allows fast evaluation of approximate matrix entries, which are needed by the classical H-matrix construction methods.

Picking the batches of points $x_i$ requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging, in which the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion  \cite{FichtnerLeeuwen15,TrampertFichtnerRitsema13}. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

%We pick batches of points $x_i$ in such that many $\phi_{x_i}$ are computed with each operator action of $\Aop$.

%We seek to approximate operators $\Aop$ that are accessible \emph{matrix-free}. That is, we have a black-box computational procedure through which we may evaluate the maps
%\begin{equation*}
%u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv
%\end{equation*} 
%for arbitrary functions $u$ and $v$, but we cannot easily compute kernel entries $\Aker(y,x)$. In applications, evaluating these maps may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.

%Product-convolution (PC) approximations can (where applicable) approximate high rank operators efficiently.

Our impulse response interpolation procedure may be categorized as a ``point spread function'' (PSF) method that is loosely based on so-called ``product-convolution'' (PC) approximations, which are approximations of an operator by weighted sums of convolution operators with spatially varying weights. PC and PSF methods have a long history dating back several decades. We note the following papers (among many others),  \cite{Adorf94,AlgerEtAl19,BigotEscandeWeiss19,NagyOleary98,FishEtAl96,EscandeWeiss12,EscandeWeiss15,EscandeWeiss22}, in which the convolution kernels are constructed from sampling impulse responses of the operator to scattered point sources. For background on PC and PSF methods, we recommend the following papers which provide well-written reviews of the topic: \cite{DenisEtAl15,EscandeWeiss17,GentileCourbinMeylan13}. While PC approximations are based on an assumption of local translation invariance, the method we propose is based on an assumption that we call ``local mean displacement invariance'' (see Figure \ref{fig:mean_displacement_invariance}). This is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). Furthermore, PC approximations have well-known issues near boundaries due to so-called ``boundary artifacts,'' caused by the boundary artificially ``cutting off'' impulse responses. We avoid this issue by temporarily excluding impulse responses from the approximation procedure whenever an attempt is made to evaluate the impulse response at a location where it is undefined.

%A single convolution operator may be high rank, so PC approximations can (where applicable) approximate high rank operators efficiently. 
%The method proposed in this paper is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of scattered points $x_i \in \Omega$ \cite{PRODCONVLIST}. 
%Often PC methods are known as ``point spread function'' (PSF) methods, particularly in optics and image processing. There, the ``operator'' $\Aop$ is a physical process of distortion and blurring of light rays passing through an optical system such as a telescope or microscope, and the local convolution kernels are shifted\footnote{The impulse responses are shifted so that each convolution kernel is centered at zero instead of the location of the point source.} impulse responses of the system to point sources of light.
%, which characterize the local distortion and blurring, and the weighting functions interpolate between these convolution kernels. 
%PSF is also used as an umbrella term to refer to other approximation methods that focus on interpolating impulse responses or impulse response features. For background on PC and PSF methods, we recommend the following papers: \cite{PRODCONVGOOD}.

%The method proposed in this paper is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses $\phi_{x_i}$ associated with a collection of scattered points $x_i \in \Omega$ \cite{PRODCONVLIST}. 
%A common choice for PC methods based on scattered impulse responses is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}. With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation.
%In our previous work, we started with a coarse grid of points $x_i$, then adaptively refined the grid in the regions where the error in the approximation was large \cite{PRODCONVMYPAPER}. However, computing each $\phi_{x_i}$ required one action of $\Aop$. Even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, were required. In this paper, we reduce the computational cost further by picking points $x_i$ in a way that allows many $\phi_{x_i}$ to be computed with each operator action of $\Aop$. This requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. The idea of estimating the supports of the functions $\phi_x$ a-priori was inspired by techniques from resolution analysis in seismic imaging, in which
%In resolution analysis, $\Aop$ is the Hessian for a seismic inverse problem, and the width of $\phi_x$ is used to estimate the minimum length scale on which features of the parameter can be inferred near the point $x$. If the width of $\phi_x$ is ten meters, for instance, then features of the parameter near $x$ can be accurately inferred from data if those features are larger than roughly ten meters in size. In \cite{RESOLUTION},
%the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion  \cite{RESOLUTION}. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Section \ref{sec:intromoments}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing at the cost of the additional constraint that $\Aop$ has a positive integral kernel.

%\subsection{Motivation: PDE constrained inverse problems}
%\label{sec:PDE_hessian_motivation}

%Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
%\begin{equation*}
%y = f(m, u(m)) + \text{noise},
%\end{equation*}
%which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
%\begin{equation}
%\label{eq:state_pde}
%0 = g(m,u).
%\end{equation}
%Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
%\begin{equation}
%\label{eq:minimization_problem}
%\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + \mathcal{R}(m),
%\end{equation}
%where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $\mathcal{R}(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, information about the Hessian of the objective function is only accessible via the action of the Hessian on vectors. In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. More accessible approximations of the Hessian are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for $m$ \cite{LAPLACEAPPROXIMATION}.

%\subsection{Background}

>>>>

% REQUIRED
\begin{abstract}
	We present a fast matrix-free method for computing hierarchical matrix (H-matrix) approximations of operators that have locally supported non-negative integral kernels. Such operators arise, for example, as Schur complements in Schur complement methods for solving partial differential equations (PDEs), Poincare-Steklov operators in domain decomposition methods, covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained inverse problems. Fast H-matrix methods can add, multiply, invert, and factorize H-matrices, compute matrix-vector products, solve linear systems, and perform other useful matrix computations. Our method computes impulse responses of the operator at a collection of scattered points, then interpolates these known impulse responses to evaluate entries of the integral kernel at other points where the impulse response is not known. Impulse responses are computed by applying the operator to a small number of Dirac combs associated with ``batches'' of point sources. By solving an ellipsoid packing problem, we choose as many points as possible per batch, while ensuring that the supports of the impulse responses within each batch do not overlap. 
	We apply the method to approximate Hessians in large-scale PDE-constrained inverse problems with highly informative data. Numerical results demonstrate that our method substantially outperforms existing state-of-the-art Hessian approximation methods which are based on low-rank approximation. We are able to form high quality approximations of high rank operators using only a small number of operator actions products.
\end{abstract}

@@

We present a fast matrix free method for approximating operators that have locally supported non-negative integral kernels. Such operators arise as Schur complements in Schur complements methods for solving partial differential equations (PDEs) \cite{TOBYPAPER}, Poincare-Steklov operators in domain decomposition methods (e.g., Dirichlet-to-Neumann maps), covariance operators in spatial statistics, blurring operators in imaging, and Hessians in distributed parameter PDE-constrained optimization and inverse problems.

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$) and let $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ be an integral operator of the following form:
\begin{equation}
	\label{eq:kernel_representation}
	(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, $\Aop u \in L^2(\Omega)'$ is the linear functional that results from applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$. We seek to approximate operators $\Aop$ that are accessible \emph{matrix-free}. That is, we have a black-box computational procedure through which we may evaluate the maps
\begin{equation*}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv
\end{equation*} 
for arbitrary functions $u$ and $v$, but we cannot easily compute kernel entries $\Aker(y,x)$. In applications, evaluating these maps may involve iteratively solving a large linear system, timestepping, or performing some other nontrivial computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representation of $\Aop$, but cannot easily compute individual matrix entries.

Our method, described in Section \ref{sec:method}, belongs to a class of methods that approximate an operator by interpolating impulse responses of the operator associated with a scattered collection of points. The \emph{impulse response} of $\Aop$ associated with the point $x$ is the function $\impulseresponse_x:\Omega \rightarrow \mathbb{R}$, 
\begin{equation}
\label{eq:impulse_response_defn}
\impulseresponse_x(y) := \Aker(y, x),
\end{equation} 
which results from keeping $x$ fixed and viewing $\Phi(y,x)$ as a function of $y$. The function $\impulseresponse_x$ is called the impulse response because straightforward analysis shows that it may be expressed as follows:
\begin{equation}
\label{eq:impulse_response_delta_action}
\impulseresponse_x = \left( \Aop \delta_x^* \right)^*,
\end{equation}
where $\delta_x$ is the the delta distribution (i.e., point source, impulse) centered at $x$, $\Aop \delta_x^* \in L^2(\Omega)'$ is an abuse of notation that denotes the result of applying $\Aop$ to the distribution $\delta_x$, and $\left( \Aop \delta_x^* \right)^* \in L^2(\Omega)$ denotes the Riesz representation of $\Aop \delta_x^*$ with respect to the $L^2$ inner product.\footnote{Recall that the Riesz representative of a functional $\sigma \in L^2(\Omega)'$ with respect to the $L^2$ inner product is the unique function $\sigma^* \in L^2(\Omega)$ such that $\sigma(v) = \left(\sigma^*,v\right)_{L^2(\Omega)}$ for all $v \in L^2(\Omega)$.} Note that while the domain of $\Aop$ is defined as $L^2(\Omega)$, which does not contain $\delta_x$, if $\Aker$ is sufficiently regular then the action of $\Aop$ may be extended to distributions. This is described in Appendix \ref{app:distributions}.

We assume:
\begin{itemize}
	\item The supports of the impulse responses $\impulseresponse_x$ are contained (or approximately contained) in localized regions.
	\item The integral kernel is non-negative in the sense that $$\Phi(y,x) \ge 0$$ for all (or nearly all) $(y,x) \in \Omega \times \Omega$.\footnote{Note that having a non-negative integral kernel is different from positive semi-definiteness. The operator $\Aop$ need not be positive semi-definite to use our method, and positive semi-definite operators need not have a non-negative integral kernel.}
\end{itemize}
Because of these assumptions, we are able to estimate the supports of the impulse responses a-priori by applying $\Aop$ to a small number of polynomial functions. We use the support estimates to compute scattered impulse responses in batches (see Figure \ref{fig:batches_intro}). This batching allows us to recover many impulse responses using a small number of operator actions $u \mapsto \Aop u$. The support estimates also allow us to interpolate the impulse responses with a new more sophisticated and accurate interpolation scheme.

\begin{figure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch1.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch12.png}
	\end{subfigure}
	\begin{subfigure}{0.32\textwidth}
		\centering
		\includegraphics[scale=0.33]{batch19.png}
	\end{subfigure}
	\caption{Batches of normalized impulse responses for the Stokes inverse problem Hessian. Left: 2nd batch. Mid: 13th batch. Right: 19th batch. Brightness scale differs between the subfigures. Black stars indicate the sample point locations at which the point source is applied. Dashed gray ellipses indicate the estimated support ellipsoids. The large circle is the boundary of $\Omega$. Early batches tend to have lots of ellipsoids near the boundary, as compared to later batches. This is because our greedy ellipsoid packing algorithm tries to pick sample points that are far away from sample points in previously chosen batches.}
	\label{fig:batches_intro}
\end{figure}

The output of our method is a hierarchical matrix (H-matrix) approximation, $\mathbf{A}_H$, of a discretization of $\Aop$. H-matrices are a matrix format in which the rows and columns of the matrix are re-ordered, then the matrix is recursively subdivided into blocks, in such a way that many off-diagonal blocks are low rank. For more details about H-matrices, see Appendix \ref{app:h_matrix}. While operator actions of $\Aop$ on vectors are typically computationally expensive, fast H-matrix methods permit us to perform matrix-vector products with $\mathbf{A}_H$ cheaply. Furthermore, one may use H-matrix methods to perform useful linear algebra operations involving $\mathbf{A}_H$ that cannot easily be done using $\Aop$, because $\Aop$ is only available matrix-free. These operations include matrix-matrix addition, matrix-matrix multiplication, matrix factorization, and matrix inversion. These H-matrix methods are fast and scale well to large problems. The ability to perform these operations permits, for example, fast solution of Newton linear systems in PDE-constrained optimization, fast sampling of ill-conditioned posterior distribution in Bayesian inverse problems, and construction of high rank surrogate models that can be used for uncertainty quantification.

\subsection{Motivation: PDE constrained inverse problems}
\label{sec:PDE_hessian_motivation}

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
	y = f(m, u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
\begin{equation}
	\label{eq:state_pde}
	0 = g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
	\label{eq:minimization_problem}
	\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + \mathcal{R}(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $\mathcal{R}(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, information about the Hessian of the objective function, $\mathcal{H}$, is only accessible via the action of $\mathcal{H}$ action on vectors. In particular, applying $\mathcal{H}$ to a vector requires solving two PDEs \cite{HESSIANACTION}. More accessible approximations of $\mathcal{H}$ are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because $\mathcal{H}^{-1}$ locally approximates the Bayesian posterior covariance for $m$.

\subsection{Background}

The most popular matrix-free operator approximation methods are based on low rank approximation \cite{CCGOPAPERS}. This includes the Lanczos method, the randomized singular value decomposition \cite{HMTRANDOM}, and CUR approximation/skeletonization \cite{CUR,MAHONEY}. However, low rank approximation is not suitable for high rank operators, and many operators of practical interest are high rank. In the context of Hessians, methods based on low rank approximation suffer from a ``data predicament''---if the data are highly informative about the unknown parameter, then the numerical rank of the data misfit term in the Hessian is large, so a large number of operator actions are required to form an accurate ``low rank'' approximation \cite{MYDISSERTATION}. 
%The ideal scenario from a scientific perspective (highly informative data) is therefore the worst case scenario from a computational perspective (large computational cost) \cite{MYDISSERTATION}.

%In general, product-convolution approximations take the form shown in \eqref{eq:product_convolution}, but $\convkernel_i$ and $w_i$ may be arbitrary functions.
Product-convolution (PC) approximations are approximations of an operator by weighted sums of convolution operators with spatially varying weights. A single convolution operator may be high rank, so PC approximations can (where applicable) approximate high rank operators efficiently. PC approximations have been used in a variety of fields going back several decades \cite{PRODCONVLIST}. Often PC methods are known as ``point spread function'' (PSF) methods, particularly in optics and image processing. There, the ``operator'' $\Aop$ is a physical process of distortion and blurring of light rays passing through an optical system such as a telescope or microscope. However, ``PSF'' is sometimes used as an umbrella term to refer to other approximation methods that focus on interpolating impulse responses or impulse response features.
%, including CUR-based low rank approximation \cite{NAGYPSFCUR}.
For background on PC and PSF methods, we recommend the following papers: \cite{PRODCONVGOOD}.

Our method is not exactly a PC approximation, but it builds upon a class of PC approximations in which the convolution kernels are shifted impulse responses of $\Aop$ associated with a collection of points $x_i \in \Omega$ \cite{PRODCONVLIST}. A popular choice for PC methods in this class is to choose the $x_i$ to be nodes in a regular grid \cite{NAGY}.
%, and interpolate the convolution operators with piecewise linear interpolation \cite{NAGY}, splines \cite{PRODCONVSPLINES}, or radial basis functions \cite{PRODCONVRBF}. 
With a regular grid, a large number of $x_i$ are typically required to achieve an accurate approximation. In our previous work, we reduced the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large \cite{PRODCONVMYPAPER}. But even with adaptive refinement, many $x_i$, and hence operator actions of $\Aop$, are required.

In this paper, 
%rather than using adaptive refinement, 
we reduce the computational cost by picking points $x_i$ in a way that allows many $\phi_{x_i}$ to be computed with each operator action of $\Aop$. This requires us to estimate the supports of the $\phi_x$, \emph{before} we compute them. We were inspired by techniques from resolution analysis in seismic imaging \cite{RESOLUTION}. In resolution analysis, $\Aop$ is the Hessian for a seismic inverse problem, and the width of $\phi_x$ is used to estimate of the minimum length scale on which features of the parameter can be inferred near the point $x$. If the width of $\phi_x$ is ten meters, then features of the parameter near $x$ can be accurately inferred from data if those features are larger than roughly ten meters in size. In \cite{RESOLUTION}, the width of $\phi_x$ is estimated to be the local autocorrelation length of the function $\Aop^T \zeta$ near $x$, where $\zeta$ is a random noise funtion. Rather than probing $\Aop^T$ with random noise functions, we use polynomial functions (see Sections \ref{sec:intromoments} and \ref{eq:mean_and_covariance_estimation}). Our method estimates the support of $\phi_x$ more accurately and reliably than random noise probing, but our method requires that $\Aop$ has a positive integral kernel, while random noise probing does not have this requirement.
%(and moreover, the seismic Hessian does not have a non-negative integral kernel).

Convolutions can be performed using the fast Fourier transform (FFT) \cite{PRODCONVFFT}, which allows PC approximations to be used as a fast surrogates for $\Aop$. Often it is necessary to perform further linear algebra operations involving $\Aop$, such as solving linear systems with $\Aop$ as the coefficient operator. Conventionally, one performs these linear algebra operations using Krylov methods, replacing $\Aop$ with its PC approximation \cite{PRODCONVKRYLOV}. The Krylov method is accelerated because performing matrix vector products with the product-convolution approximation is typically faster than performing operator actions with $\Aop$. However, there are several problems with this approach. First, the number of Krylov iterations can be large, particularly when $\Aop$ is high-rank. Second, complicated geometries are rarely discretized using regular grids, so using the FFT to perform convolutions requires transferring discretized functions back and forth between the irregular mesh and a regular grid. Third, the computational cost grows with the number of convolution kernels used. 
%If a large number of convolution kernels are used, this approach can be slow.
%Furthermore, in big-data inverse problems the required number of Krylov iterations is large \cite{MYDISSERTATION,OTHERS}. 
Thus, 
%after constructing the product-convolution approximation, 
it is desirable to convert the PC approximation to other matrix formats that are amenable to faster linear algebra operations. Wavelet compression methods have been used for this purpose \cite{PRODCONVWAVELETS}, because operators that are amenable to PC approximation are often sparse in the wavelet domain. Convolution-product approximations (transposes of PC approximations) are equivalent to low rank approximation of the pseudo-differential symbol of the operator, so one could perform linear algebra operations involving PC approximations using fast methods for pseudo-differential operator symbol calculus \cite{YINGDEMANET}. Other matrix-free operator approximation methods based on wavelets \cite{HERRMANN} and pseudo-differential operators \cite{DEMANET} do not use PC approximation. Here, we follow our previous work \cite{PRODCONVMYPAPER}, in which we converted PC approximations into H-matrix format.

Since our end goal is constructing an H-matrix, we are freed from strict adherence to the PC format. This allows us to modify the PC approach to address well-known deficiencies. PC approximations are based on an assumption of local translation invariance \cite{LOCATRANSLATIONINVARIANCE}, while our method is based on an assumption that we call ``normalized local mean displacement invariance,'' which is more general than local translation invariance, and includes both low rank approximation and local translation invariance as special cases (see Section \ref{sec:local_mean_displacement_invariance}). PC approximations have well-known issues near boundaries called boundary artifacts, which are caused by the boundary artificially ``cutting off'' impulse responses. 
%One must extend the impulse responses outside of the domain where they are undefined, but
%. Various extension techniques have been used, such as extension by zero, and reflection across the boundary, yet 
%no known extension technique is fully satisfactory \cite{BOUNDARYETENSION}. 
In our method we do not have an issue with boundary artifacts because we temporarily exclude impulse responses from the approximation procedure whenever the approximation procedure attempts to evaluate the impulse response at a location where it is undefined. In exchange for these benefits, our approximation cannot be written as a weighted sum of convolution operators, so we cannot apply our approximation to vectors via the FFT. This is fine because our approximation is an H-matrix, which can be applied to vectors even faster.

%MOMENTS PERMIT IMPROVED INTERPOLATION. IMPROVED INTERPOLATION REQUIRES H-MATRIX. H-MATRIX WITH IMPROVED INTERPOLATION ALSO SOLVES BOUNDARY ARTIFACT PROBLEM. ALSO DOESNT REQUIRE REGULAR GRID. ALSO SCALABLE TO LARGE NUMBER OF IMPULSE RESPONSES. FFT HAS FAST MATVEC BUT NOT OTHER OPTIONS.
